---
title: "Styles"
format: html
execute:
  echo: true
  warning: false
  message: false
  cache: false
vignette: >
  %\VignetteIndexEntry{a05_style}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
# Setup chunk
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7.2,
  fig.height = 5
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# Introduction

In this vignette, we explain how to customise the visualisation of tables and plots. The vignette reviews the structure of the `.yml` files that define styles, and demonstrates how to create and apply custom styles. It also shows how to style tables and plots programmatically, without the need to create a `.yml` file.

```{r, message=FALSE, warning=FALSE}
library(visOmopResults)
library(here)
library(gt)
library(ggplot2)
library(dplyr)
library(officer)
```

The package currently includes two built-in styles for tables and plots. Styles are defined using `.yml` files. To list the available styles, use:

```{r}
tableStyle()
plotStyle()
```

# Branding styles using `.yml`

The package contains two built-in styles: `"default"` and `"darwin"`. The `.yml` files for these styles can be found [here](https://github.com/darwin-eu/visOmopResults/tree/main/inst/brand).

## `.yml` structure 

We use the `"darwin"` style as an example. The code chunk below shows the structure of its `.yml` file:

```{r, echo=FALSE, results='asis'}
library(yaml)
theme <- read_yaml(file = here("inst/brand/darwin.yml"))
cat("```yaml\n", as.yaml(theme), "```", sep = "")
```

The `.yml` structure can be divided into four main sections:

- **Color:** Defines the color palette and the default `background` and `foreground` colors used when a plot/table section does not override them.
- **Typography:** Defines default font families and sizes for base text, plots, and tables (these can be overridden in the plot/table sections).
- **Plot:** Plot-specific settings such as background color, facet header color, header text color and bold, grid color, axis color, border color, and legend position. Font settings are taken from the typography section unless overridden here.
- **Table:** Table-specific settings. You can set an overall `border-color` and `border-width`, or override settings per table section. Table sections include: `header`, `header-name`, `header-level`, `column-name`, `group-label`, `title`, `subtitle`, and `body`. For each section you can set properties such as `background-color`, `text-color`, `text-bold`, `align`, `font-size`, `border-color`, and `border-width`.

## Style hierarchy

Each plot and table element follows a style hierarchy. If a value isn’t specified at the most specific level, it inherits from higher-level entries; if none are defined, the default `ggplot2` (for plots) or the default for the specific table type is used. The table below shows the priority order for common plot and table options.

```{r, echo=FALSE}
tibble(
  "type" = "Plot",
  "Part" = c("Background color", "Header (facet) color", "Header (facet) text color", "Header (facet) text bold", "Border color", "Grid color", "Axis color", "Legend position", "Font family", "Font size"),
  "Option 1" = c("defaults:visOmopResults:plot:background-color", "defaults:visOmopResults:plot:header-color", "defaults:visOmopResults:plot:header-text-color", "defaults:visOmopResults:plot:header-text-bold", "defaults:visOmopResults:plot:border-color", "defaults:visOmopResults:plot:grid-major-color", "defaults:visOmopResults:plot:axis-color", "defaults:visOmopResults:plot:legend-position", "defaults:visOmopResults:plot:font_family", "defaults:visOmopResults:plot:font_size"),
  "Option 2" = c("color:background", "color:foreground", "-", "color:foreground", "-", "color:foreground", "-", "-",  "typography:base:family", "defaults:visOmopResults:plot:font_size"),
  "Option 3" = c(rep("-", 9), "typography:base:size")
) |> 
  bind_rows(
    tibble(
      "type" = "Table section",
      "Part" = c("Background color", "Text bold", "Text color", "Text align", "Font size", "Font family", "Border color", "Border width"),
      "Option 1" = c("defaults:visOmopResults:table:[section_name]:background-color", "defaults:visOmopResults:table:[section_name]:text-bold", "defaults:visOmopResults:table:[section_name]:text-color", "defaults:visOmopResults:table:[section_name]:align", "defaults:visOmopResults:table:[section_name]:font-size", "defaults:visOmopResults:table:[section_name]:font-family", "defaults:visOmopResults:table:[section_name]:border-color", "defaults:visOmopResults:table:[section_name]:border-width"),
      "Option 2" = c("color:background", "-", "-", "-", "defaults:visOmopResults:table:font-size", "defaults:visOmopResults:table:font_family", "defaults:visOmopResults:table:border-color", "defaults:visOmopResults:table:border-width"),
      "Option 3" = c("-", "-", "-", "-", "defaults:visOmopResults:typography:base:size", "typography:base:family", "-", "-")
    )
  ) |>
  visTable(groupColumn = "type")
```

In the examples above the YML path is represented with colon separators. For example, `plot:background-color` refers to the `background-color` key inside the `plot` section.

## Applying styles to tables and plots

The table-formatting functions (`visTable()`, `visOmopTable()`, and `formatTable()`) and plot functions accept a `style` argument. The `style` argument can be:

- the name of a built-in style (e.g. `"darwin"`), or  
- the path to a user `.yml` file that defines a custom style, or  
- a programmatic list that mirrors the `.yml` structure (only tables - see next section).

Example: apply the built-in `"darwin"` style to a plot:

```{r}
result <- mockSummarisedResult() |> 
  filter(variable_name == "age")

barPlot(
  result = result,
  x = "cohort_name",
  y = "mean",
  facet = c("age_group", "sex"),
  colour = "sex",
  style = "darwin"
)
```

Example: use a custom `.yml` file (path provided):

```{r, eval=FALSE}
barPlot(
  result = result,
  x = "cohort_name",
  y = "mean",
  facet = c("age_group", "sex"),
  colour = "sex",
  style = here("MyStyleFolder", "MyStyle.yml")
)
```

## Use of `_brand.yml`

If `style = NULL` and no global options are provided (via `setGlobalPlotOptions()` or `setGlobalTableOptions()`), the built-in "default" style is used. However, if a `_brand.yml` file is present in the project directory, that file’s style will be used.

# Alternative style customisation

You can customise styles programmatically without creating a `.yml` file by passing a named list to the `style` argument. The list should follow the same table section structure as the `.yml`.

## Tables 

Below is an example that sets table section styles for `gt`. 

```{r}
result |>
  visOmopTable(
    estimateName = c("Mean (SD)" = "<mean> (<sd>)"),
    groupColumn = "cohort_name",
    header = c("This is an overall header", "sex"),
    type = "gt",
    style = list(
      header = list(
        cell_text(weight = "bold"), 
        cell_fill(color = "red")
      ),
      header_name = list(
        cell_text(weight = "bold"), 
        cell_fill(color = "orange")
      ),
      header_level = list(
        cell_text(weight = "bold"), 
        cell_fill(color = "yellow")
      ),
      column_name = list(
        cell_text(weight = "bold")
      ),
      group_label = list(
        cell_fill(color = "blue"),
        cell_text(color = "white", weight = "bold")
      ),
      title = list(
        cell_text(size = 20, weight = "bold")
      ),
      subtitle = list(
        cell_text(size = 15)
      ),
      body = list(
        cell_text(color = "red")
      )
    ),
    .options = list(
      title = "My formatted table!",
      subtitle = "Created with the `visOmopResults` R package.",
      groupAsColumn = FALSE,
      groupOrder = c("cohort2", "cohort1")
    )
  )
```


Note that style objects differ across table engines, so the code must be adapted to the engine you use.

For `flextable`, styling objects come from the `officer` package. The structure is similar, but the style objects differ:

```{r}
result |>
  visOmopTable(
    estimateName = c("Mean (SD)" = "<mean> (<sd>)"),
    groupColumn = "cohort_name",
    header = c("This is an overall header", "sex"),
    type = "flextable",
    style = list(
      header = list(
        cell = fp_cell(background.color = "red"),
        text = fp_text(bold = TRUE)
      ),
      header_level = list(
        cell = fp_cell(background.color = "orange"),
        text = fp_text(bold = TRUE)
      ),
      header_name = list(
        cell = fp_cell(background.color = "yellow"),
        text = fp_text(bold = TRUE)
      ),
      column_name = list(
        text = fp_text(bold = TRUE)
      ),
      group_label = list(
        cell = fp_cell(background.color = "blue"),
        text = fp_text(bold = TRUE, color = "white")
      ),
      title = list(
        text = fp_text(bold = TRUE, font.size = 20)
      ),
      subtitle = list(
        text = fp_text(font.size = 15)
      ),
      body = list(
        text = fp_text(color = "red")
      )
    ),
    .options = list(
      title = "My formatted table!",
      subtitle = "Created with the `visOmopResults` R package.",
      groupAsColumn = FALSE,
      groupOrder = c("cohort2", "cohort1")
    )
  )
```

## Plots

Plot helpers return `ggplot2` objects, so you can further modify them using `+` and regular `ggplot2` calls:

```{r}
library(ggplot2)

barPlot(
  result = result,
  x = "cohort_name",
  y = "mean",
  facet = c("age_group", "sex"),
  colour = "sex"
) +
  theme(
    strip.background = element_rect(fill = "#ffeb99", colour = "#ffcc00"),
    legend.position = "top",
    panel.grid.major = element_line(color = "transparent", linewidth = 0.25)
  ) +
  scale_color_manual(values = c("black", "black", "black")) +
  scale_fill_manual(values = c("#999999", "#E69F00", "#56B4E9"))
```

# Using non-registered font families in ggplot2

To use a specific font family in ggplot2, the font must be:

1. **Installed** in the operating system, and

2. **Available** to R’s graphics device (registered, in the case of Windows).

Below is an example using the Calibri font.

## 1. Install the font in the system

On both **macOS** and **Windows**, install the .ttf file by double-clicking it and clicking Install.

Example source: https://www.freefontdownload.org/en/calibri.font

After installing new system fonts, **restart R** or **RStudio** so the font registry is refreshed.

## 2. Register the font 

On **macOS**, most system fonts are automatically available to R’s Quartz graphics device (no need to register).

On **Windows**, however, the base graphics device does not automatically expose all installed system fonts.
You must register a font before ggplot2 can use it. This can be done as follows:

```{r, eval=FALSE}
windowsFonts(Calibri = windowsFont("Calibri"))
```

* The `visOmopResults` package automatically registers any installed font when needed, so users generally do not have to run this manually.

## 3. Create plots with styles that use the font

You can specify the font family in your YAML configuration, or directly in `theme()` using `element_text()`.
Below is an example using the "darwin" plot style, which will use "Calibri" when available, otherwise falling back to "sans":

```{r}
barPlot(
  result = result,
  x = "cohort_name",
  y = "mean",
  facet = c("age_group", "sex"),
  colour = "sex",
  style = "darwin"
)
```

## Important notes
- After installing system fonts, restart R/RStudio so R can detect them.

- On Windows, font registrations done with `windowsFonts()` last only for the current R session and revert after restarting.

- For font detection across platforms, `visOmopResults` uses the `systemfonts` package and registers fonts on Windows when needed.

# Final remarks

The `.yml` customisation system allows you to control most aspects of the visual appearance of your tables and plots. To learn more about `brand.yml` and how it interacts with other elements such as Shiny apps and Quarto/R Markdown documents, refer to <https://posit-dev.github.io/brand-yml/>.
