---
title: "format functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{format functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

The **format** set of functions can be combined to format a summarised_result object (see R package *omopgenerics*) into a nice *tibble*, *flextable* or *gt* table to display in reports and shiny apps. 

The functions `formatEstimateValue()`, `formatEstimateName()`, `formatHeader()`, and lastly `gtTable()` or `fxTable()` can be implemented in a pipeline to obtain the desired formatted table. Otherwise, these steps are implemented in the function `visOmopTable()`. 

## Formatting functions in a pipeline

First, we load the relevant libraries and generate a summarised_result with mock results.

```{r setup}
library(visOmopResults)
library(dplyr)
mock_sr <- mockSummarisedResult()
mock_sr |> glimpse()
```

### 1. formatEstimateValue()
This function provides tools to format the estimate_value column: choose *number of decimals* to display for each estimate_type or estimate_name (`decimals`), and change  change the *decimal and thousand/million separator* mark (`decimalMark` and `bigMark` respectively)

By default, decimals of integer values are set to 0, numeric to 2, percentage to 1, and proportion to 3. The defaulted decimal mark is "." while the thousand/million separator is ",".

```{r}
mock_sr <- mock_sr |> formatEstimateValue()
mock_sr |> glimpse()
```

If we want the same number of decimals for all the estimates, instead of a named vector the `decimal` argument can be a numeric pointing the number of decimals.

### 2. formatEstimateName()
We can *customise estimate display* by changing the estimate name and joining different estimates in the same row. For instance, we can display counts and percentages together as "N (%)".
 
The `estimateNameFormat` is where combinations can be specified. Values from *estimate_name*'s column should be specified between <...>. The new *estimate_name* will be the names of this vector, or the value itself when it is not named. 

```{r}
mock_sr <- mock_sr |> 
  formatEstimateName(
    estimateNameFormat = c(
      "N (%)" = "<count> (<percentage>%)", 
      "N" = "<count>",
      "Mean (SD)" = "<mean> (<sd>)"
    ),
    keepNotFormatted = FALSE,
    useFormatOrder = FALSE
  )
mock_sr |> glimpse()
```

Additional input arguments are `keepNotFormatted` to specify whether not formatted rows should be returned or dropped.
If `useFormatOrder` is TRUE, estimates will be presented in the order given in `estimateNameFormat`, if FALSE, the original order will be kept (where there is more than one estimate, the new position will be the first of the estimates being merged).


### 3. formatHeader()

This function helps to create a nice header for a *flextable* or *gt* table. For instance, instead of having a column specifying the group name and level, we might one them in the header. 

**Header keys**: there are 3 different types of headers, identified with the keys "header", "header_name", and "header_level".

- **header:** Header labels passed to `header` but which are not part of the input table (are not columns names or values).
- **header_name:** Header labels that come from column names in the input table.
- **header_level:** Header labels that come from values of selected columns in the input table.

For instance, we might want to pivot by "group_level" and have an upper header called "Names of the cohorts". To do that we would proceed as follows:

```{r}
mock_sr |>
  formatHeader(
    header = c("Names of the cohorts", "group_level"),
    delim = "\n",
    includeHeaderName = TRUE,
    includeHeaderKey = TRUE
  ) |>
  glimpse()
```

The label keys can be removed with `includeHeaderKey` set to FALSE. However, having these keys in our header will allow to style separately the different header types in the next step (`fxTable` and `gtTable`).
If we wish to remove the header pointing to the column name (*group_level*), we can set `includeHeaderName` to FALSE. 

Continuing with our example, we want to pivot by strata (name and level), but, we do not want the column names (that is, *strata_name* and *strata_level*) to appear in the header:

```{r}
mock_sr <- mock_sr |>
  mutate(across(c("strata_name", "strata_level"), ~ gsub("&&&", "and", .x))) |>
  formatHeader(
    header = c("Stratifications", "strata_name", "strata_level"),
    delim = "\n",
    includeHeaderName = FALSE,
    includeHeaderKey = TRUE
  ) 

mock_sr |> glimpse()
```

Notice, how we substitute the keyWord "&&&" to "and", to get a nice header.

 
### 4. gtTable() and fxTable()

Finally, we have the functions `gtTable` and `fxTable` which will transform our tibble into a *gt* or *flextable* object respectively. These functions provide several tools to personalise the formatted table. 

#### gtTable()
Let's start by manipulating the dataframe to have the columns that we want to display, and then use `gtTable` with default values:

```{r}
# first we select the columns we want:
mock_sr <- mock_sr |>
  splitGroup() |>
  select(!all_of(c(
    "cdm_name", "estimate_type", "result_id","additional_name", 
    "additional_level"
  ))) 
mock_sr |>  gtTable()
```

Now, we want to group results by "cohort_name". More specifically we want to have a row which the name of each cohort before the results of that cohort, and that *cohort1* comes before *cohort2*. Additionally, we want to merge those rows what contain the same information for all the columns. To get this table we will use `gtTable` as follows:

```{r}
mock_sr |>  
  gtTable(
    groupNameCol = "cohort_name",
    groupNameAsColumn = FALSE,
    groupOrder = c("cohort1", "cohort2"),
    colsToMergeRows = "all_columns"
  )
```


We might also want to modify the default style of the table. For instance, we might want to highlight the cohort_name labels with a blue background, have the body text in red, and use a combination of orange and yellow for the header. We can do it with the `style` argument:

```{r}
mock_sr |>  
  gtTable(
    style = list(
      "header" = list(gt::cell_text(weight = "bold"), 
                      gt::cell_fill(color = "orange")),
      "header_level" = list(gt::cell_text(weight = "bold"), 
                      gt::cell_fill(color = "yellow")),
      "column_name" = gt::cell_text(weight = "bold"),
      "group_label" = list(gt::cell_fill(color = "blue"),
                           gt::cell_text(color = "white", weight = "bold")),
      "body" = gt::cell_text(color = "red")
    ),
    groupNameCol = "cohort_name",
    groupNameAsColumn = FALSE,
    groupOrder = c("cohort1", "cohort2"),
    colsToMergeRows = "all_columns"
  )
```


#### fxTable()
To obtain a similar result but with a *flextable* object, we can use `fxTable` with the same arguments as before, however, `style` must be adapted to use the *officer* package since it is the accepted by *flextable*.

```{r}
mock_sr |>  
  fxTable(
    style = list(
      "header" = list(
        "cell" = officer::fp_cell(background.color = "orange"),
        "text" = officer::fp_text(bold = TRUE)),
      "header_level" = list(
        "cell" = officer::fp_cell(background.color = "yellow"),
        "text" = officer::fp_text(bold = TRUE)),
      "column_name" = list("text" = officer::fp_text(bold = TRUE)),
      "group_label" = list(
        "cell" = officer::fp_cell(background.color = "blue"),
        "text" = officer::fp_text(bold = TRUE, color = "white")),
      "body" = list("text" = officer::fp_text(color = "red"))
    ),
    groupNameCol = "cohort_name",
    groupNameAsColumn = FALSE,
    groupOrder = c("cohort1", "cohort2"),
    colsToMergeRows = "all_columns"
  )
```


## visOmopTable(): all at once

This function wraps the tools seen so far to format a summarised result at once. While it is convenient to have it all in one function, the level of table manipulation allowed is less than with the pipeline.

First, we create a new mock summarised result:
```{r }
mock_sr2 <- mockSummarisedResult()
```

For instance, if we want to have a flextable with strata as header, we will write "strata" in `header` instead of *strata_name* and *strata_level*. Same as before, we can input header keys (e.g. "Stratifications" in the example below).

The `split` argument points to which name-level columns to split (group, strata and/or additional). Refer to the vignette split and unite for further information on this functionality.

```{r}
mock_sr2 |> visOmopTable(
  formatEstimateName = c("N%" = "<count> (<percentage>)",
                         "N" = "<count>",
                         "Mean (SD)" = "<mean> (<sd>)"),
  header = c("Stratifications", "strata"),
  split = c("group","additional")
)
```
By default, it returns a gt table, but it can be changed to "flextable" and "tibble" in the `type` argument.

The `groupColumn` can be used to create groups in the table body. For instance, looking at the previous table, we might want to group by cohort name:

```{r}
mock_sr2 |> visOmopTable(
  formatEstimateName = c("N%" = "<count> (<percentage>)",
                         "N" = "<count>",
                         "Mean (SD)" = "<mean> (<sd>)"),
  header = c("Stratifications", "strata"),
  split = c("group","additional"),
  groupColumn = "cohort_name"
)
```

Additional options can be specified in the `.options` argument. These options are taken from the other format functions and gt/fx table functions seen before. To see a list of allowed arguments and their default values use `optionsVisOmopTable()`:

``` {r}
optionsVisOmopTable()
```
