---
title: "Plotting tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=7, 
  fig.height=5
)
```

**visOmopResults** offer some plotting tools that can be very useful. These 
tools are useful to plot either `<summarised_result>` dataset or a usual
`<data.frame>`.

```{r setup}
library(visOmopResults)
```

## Plotting with a <summarised_result>

For the purpose of this vignette we will turn summarise the dataset penguins 
from the palmerpenguins package using `PatientProfiles::summariseResult()`

```{r}
library(PatientProfiles)
library(palmerpenguins)
library(dplyr, warn.conflicts = FALSE)

summariseIsland <- function(island) {
  penguins |>
    filter(island == island) |>
    summariseResult(
      group = "species",
      includeOverallGroup = TRUE,
      strata = list("year", "sex", c("year", "sex")),
      variables = c(
        "bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g", 
        "sex"),
      estimates = c(
        "median", "q25", "q75", "min", "max", "count_missing", "count", 
        "percentage", "density")
    ) |>
    suppressMessages() |>
    mutate(cdm_name = island)
}

penguinsSummary <- bind(
  summariseIsland("Torgersen"), 
  summariseIsland("Biscoe"), 
  summariseIsland("Dream")
)
```

### Plotting principles

Although the input of a visOmopResult plot function can be a <summarised_result>
when referring to inputs, we will use the columns shown in its tidy format:

```{r}
tidy(penguinsSummary) |>
  glimpse()
```

We should always subset the <summarised_result> to the variable_name of interest
before calling the plotting functions. It is not advised to combine results of
different variables if they don't share common estimates that we want to plot as
in the tidy form NAs will be created.

### Scatter plot

```{r}

```


### Bar plot

```{r}

```


### Box plot

Let's create some box plots of their body mass:

```{r}
penguinsSummary |>
  filter(variable_name == "body_mass_g") |>
  plotBoxplot(x = "year", facet = c("cdm_name", "species"), colour = "sex")
```

We can specify how we want to facet using a formula:
```{r}
penguinsSummary |>
  filter(variable_name == "body_mass_g") |>
  plotBoxplot(x = "year", facet = cdm_name ~ species, colour = "sex")
```

Arguments before `~` specify the rows, and arguments after the columns, see 
another example faceting also by sex in the columns. And in this case colouring 
by year:
```{r}
penguinsSummary |>
  filter(
    group_level != "overall",
    strata_name == "year &&& sex",
    !grepl("NA", strata_level),
    variable_name == "body_mass_g") |>
  plotBoxplot(facet = cdm_name ~ species + sex, colour = "year")
```

Note that as we didnt specify x there is no levels in the x axis, but box plots
are produced anyway.

## Plotting with a <data.frame>

Plotting functions can also be used with a normal <data.frame>. In this case we 
will use the tidy format of `penguinsSummary`.

```{r}
penguinsTidy <- penguinsSummary |>
  filter(!estimate_name %in% c("density_x", "density_y")) |> # remove density for simplicity
  tidy()
penguinsTidy |> glimpse()
```

That as it can be seen is a normal data.frame:
```{r}
penguinsTidy |> class()
```

In fact, the tidy format makes it easy to filter the group, strata and 
additional columns as they are not combined in the name-level format. But it is
also quite useful to filter to the variables of interest, before applying the
tidy format.
```{r}

```

## Custom plotting

The tidy format is very useful to apply any other custom ggplot2 function that 
we may be interested on:
```{r}
library(ggplot2)
penguinsSummary |>
  filter(variable_name == "number records") |>
  tidy() |>
  ggplot(aes(x = year, y = sex, count = count)) +
  geom_bin_2d() +
  facet_grid(cdm_name ~ species)
```

## Combine with `ggplot2`

The plotting functions are a wrapper around the ggplot2 package, outputs of 
the plotting functions can be later customised with ggplot2 and similar tools. 
For example we can use `ggplot2::labs()` to change the labels and 
`ggplot2::theme()` to move the location of the legend.

```{r}
penguinsSummary |>
  filter(
    group_level != "overall",
    strata_name == "year &&& sex",
    !grepl("NA", strata_level),
    variable_name == "body_mass_g") |>
  plotBoxplot(facet = cdm_name ~ species + sex, colour = "year") +
  ylim(c(0, 6000)) +
  labs(x = "My custom x label") +
  theme(legend.position = "top")
```

You can also use `ggplot2::ggsave()` to later save one of this plots into '.png'
file.

```{r, eval=FALSE}
ggsave(
  "figure8.png", plot = last_plot(), device = "png", width = 15, height = 12, 
  units = "cm", dpi = 300)
```

## Combine with `plotly`

Although the package currently does not provide any plotly functionality
ggplots can be easily converted to `<plotly>` ones using the function
`plotly::ggplotly()`. This can make the interactivity of some plots better:

```{r}
```
