---
title: "split and unite functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{split and unite functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

**Split** and **unite** are complementary functions to manipulate dataframes in R. They work with summarised_results objects (see R package *omopgenerics*), but they can also support R dataframes from other classes.

## summarised_result

First, let's load relevant libraries and generate a mock summarised_result object to use in the following examples.
```{r setup}
library(visOmopResults)
library(dplyr)
mock_sr <- mockSummarisedResult()
mock_sr |> glimpse()
```


A summarised_result contains 3 types of name-level paired columns which are targeted by the set of unite and split functions. These are the **group** columns which typically can contain information about cohorts, **strata** columns which have data on stratification for each group, and finally the **additional** columns which include further information not covered by group and strata.

## Split functions
The idea of the split functions is to pivot the "name" (e.g. *group_name*) column to split each value of that column into a column in the dataframe, which values are taken by the "level" (e.g. *group_level*) column.

### splitGroup(), splitStrata(), and splitAdditional()

For instance, the `splitGroup` function will target the *group_name-group_level* columns as seen below. 

```{r}
mock_sr |> splitGroup() |> glimpse()
```

Similar to `splitStrata`, the functions `splitGroup` will split *group_name* and *group_level* columns, while `splitAdditional` will split the additional name-level pair. Finally, the function `splitAll` will split group, strata, and additional at once. Note that after using `splitStrata` on our summarised_result object, we do no longer have a *strata_name-strata_level* pair, instead we have two new columns corresponding to the stratifications, *age_group* and *sex*.

```{r}
mock_sr |> splitStrata() |> glimpse()
mock_sr |> splitAdditional() |> glimpse()
mock_sr |> splitAll() |> glimpse()
```

#### !! Keyword: &&&
Looking at the results below, observe how the splitting was not only done by values in the "name" column, but also among values containing the key word "**&&&**". That is, "sex &&& age_group" was splitted into *sex* and *age_group* columns, instead of generating a column called "*sex &&& age_group*".

### splitNameLevel()
The function `splitNameLevel` provides a more tailored splitting of the dataframe. This function can take any dataframe with no restrictions to the naming of the name-level pair columns, since these can be specified in the `name` and `level` arguments. 

For instance let's use it in the following table:
```{r}
data_to_split <- tibble(
  denominator = "general_population",
  outcome = "stroke",
  input_arguments = c("wash_out &&& previous_observation"),
  input_arguments_values = c("60 &&& 180")
)
data_to_split 
```


```{r}
data_to_split |>
  splitNameLevel(
    name = "input_arguments",
    level = "input_arguments_values"
  )
```

The function `splitNameLevel`, in additionally to the argument `overall` previously seen, has the argument `keep` to set whether we want to keep the columns before the splitting.



## Unite functions

The unite functions are the complementary to the split ones. These are meant to generate name-level pair columns from targeted columns within a dataframe. 

### uniteGroup(), uniteStrata(), and uniteAdditional()
To work with summarised_result objects, we have the `uniteGroup`, `uniteStrata`, and `uniteAdditional` functions which will generate the group, strata, and additional name-level columns respectively from a given set of columns. For instance, in the following example we want to create the *group_name* and *group_level* columns:

```{r}
to_unite_group <- tibble(
  denominator_cohort_name = c("general_population", "older_than_60", "younger_than_60"),
  outcome_cohort_name = c("stroke", "stroke", "stroke")
)

to_unite_group |>
  uniteGroup(cols = c("denominator_cohort_name", "outcome_cohort_name"))
```

A part from the columns to unite argument (`cols`), there is the argument `ignore`, by default: `ignore = c(NA, "overall")`. This means that, levels within ignore will be ignored. For example if in this case we do not ignore them we will obtain the NA as output:

```{r}
to_unite_strata <- tibble(
    age = c(NA, ">40", "<=40", NA, NA, NA, NA, NA, ">40", "<=40"),
    sex = c(NA, NA, NA, "F", "M", NA, NA, NA, "F", "M"),
    region = c(NA, NA, NA, NA, NA, "North", "South", "Center", NA, NA)
  )

to_unite_strata |>
  uniteStrata(cols = c("age", "sex", "region"),
              ignore = character())
```

By default (`ignore = c(NA, "overall")`) we obtain an output where only names and levels of non-NA values are returned, and from those rows where all values are NA it uses "overall".
```{r}
to_unite_strata |>
  uniteStrata(cols = c("age", "sex", "region"))
```

### uniteNameLevel()

Lastly, the function `uniteNameLevel`, idem to `splitNameLevel`, provides more flexibility on the name-level column naming, in addition of the `keep` argument (FALSE by default) to choose whether to keep the targeted columns. For instance, if we repeat the previous example with `keep` set to TRUE we would obtain the following output:

```{r}
to_unite_strata |>
  uniteNameLevel(cols = c("age", "sex", "region"),
                 name = "name",
                 level = "level",
                 keep = TRUE)
```
