---
title: "tidy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```


The package provides functionalities to tidy a summarised result to obtain a dataframe with which is easier to do subsequent calculations. 

In this line, the `split` functions, described in **split and unite functions** allow to interact with name-level columns. 

For the estimates, we have the `pivotEstimates` function, and for the settings `pivotSettings`. Finally the `tidy` method accommodates the split and pivot functionalities in the same function.

## Estimates
First, let's load relevant libraries and create a mock summarised result table.
```{r }
library(visOmopResults)
library(dplyr)
result <- mockSummarisedResult()
result |> glimpse()
```

The function `pivotEstimates` adds columns containing the estimates values for each combination of columns in `pivotEstimatesBy`. For instance, in the following example we use the columns *variable_name*, *variable_level*, and *estimate_name* to pivot the estimates.

```{r }
result |> 
  pivotEstimates(pivotEstimatesBy = c("variable_name", "variable_level", "estimate_name")) |>
  glimpse()
```

The argument `nameStyle` is to customise the names of the new columns. It uses the glue package syntax. For instance:

```{r }
result |> 
  pivotEstimates(pivotEstimatesBy = "estimate_name",
                 nameStyle = "{toupper(estimate_name)}") |>
  glimpse()
```

## Settings
The function `pivotSettings` adds a new column for each of the settings in the summarised result, if any:
```{r}
mockSummarisedResult(settings = TRUE) |>
  pivotSettings() |>
  glimpse()
```

The function `appendSettings` is the inverse functionality to pivot. This function will append columns corresponding to settings to the end summarised result:

```{r}
table <- mockSummarisedResult() |>
  mutate(mockSummarisedResult = TRUE, vignette = "tidy")

result <- table |> appendSettings(colsSettings = c("mockSummarisedResult", "vignette"))

result |> filter(variable_name == "settings") |> glimpse()
```

## Tidy
Finally, the method `tidy` incorporates the splitting pf name-level columns and pivotting of estimates and settings. By default, it splits group, strata and additional, pivots estimates by the columns "estimate_name" and also pivots the settings. 

```{r}
result <- mockSummarisedResult(settings = TRUE)

result |> 
  tidy() |> 
  glimpse()
```

Which column pairs to split can be customised with the *split* arguments, while `pivotEstimatesBy` and `nameStyle` are for pivotting estimates. If `pivotEstimatesBy` is `NULL` or `character()`, estimates will not be modified. Settings will always be pivotted if present. 
